Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-10-20 paper/type-system.tex
Edits.

The expression language has been equipped with lambda abstraction
($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
expressions to define and apply anonymous functions. Modulo the angle
braces, these expressions are uncurried variants of corresponding
expressions from System F. Note that FGJ formalism does not include
higher-order functions, hence the syntactic class (\C{T}) of FGJ types
needs to be extended with an (uncurried) arrow type.

Informally, \FB class definitions are extensions of FGJ class
definitions with regions. A class definition in FGJ can be lifted to
its \FB definition by (a). lifting all core types in the body of the
class (i.e, fields, constructors and methods) to their
region-annotated versions, (b). parameterizing the class over
region variables that occur free in the region-annotated body of the
class, and (c). recording any outlives constraints on the region
parameters of the class. The later two steps effectively make the
class region polymorphic (cf. \S\ref{sec:alloc-ctxt}: qualified region
polymorphism). Note that in \FB, regions, and their static identifiers
($\rgn$) are confined to the expression language. Consequently, the
region annotated types of class fields, constructors and methods do
not refer to identifiers of concrete regions, but rather their place
holders called \emph{region variables} ($\rho$). Informally, region
variables ($\rho$) are to region identifiers ($\rgn$) as type
variables ($a$) are to types ($T$).

@@

\FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
$\RgnZ\inang{\rgn}$. However, $\rgn$ cannot be any region. Recall
that $\RgnZ$ objects have special semantics in \name - they act as
handlers to transferable regions. Constructing a new $\RgnZ$ object
entails the creation of a new transferable region, and it is in this
region that the new object is allocated in. It follows that $\rgn$ in
$\RgnZ\inang{\rgn}$ should be the static identifier of the new
transferable region. But, static identifiers for transferable
regions are introduced only when such regions are opened for
allocation via \C{open} expression. What, then, should be the region
type of a \C{new Region} expression?

@@

Existentially quantifying the allocation region in the type of $\RgnZ$
objects is also useful to assign appropriate types to data structures
of transferable regions. For example, a list of transferable regions
has the type $\C{List}\inang{??}$ in \FB.

A method definitions ($d$) in \fbname can be region polymorphic with
respect to (a). its allocation context (Sec.~\ref{sec:alloc-ctxt}),
and (b). the regions occuring in the region types of its arguments.
Region parameters on the methods, like those on classes, are
accompanied by constraints ($\phi$) capturing the conditions that the
parameters need to satisfy for the method to be considered well-formed
(Fig.~\ref{fig:fb-morewfrules}). Allocation context ($\rho^a$ or
$\rho^a_m$) is the first and inevitable region parameter of every
method in \FB. If a method is not polymorphic with respect to its
allocation context (for example, if its allocation context needs to be
same as the allocation region of its class), then the monomorphism
needs to be encoded as an explicit equality constraint in $\phi$.

\fbname supports first-class functions via lambda expressions and
arrow types. A lambda expression defines a region-polymorphic
multi-argument function closure parameterized over function's
allocation context parameter. The arrow type of a function closure is
prenex-quantified over its constrained region arguments
($\inang{\rho^a\rhobar \,|\, \phi}$). Since closures can escape the
context in which they are created, it is important to keep track of
the regions where closures are allocated in order to prevent unsafe
dereferences. The $\rgn^a$ annotation above the arrow in the arrow
type denotes the allocation region of the corresponding closure. Note
that it is important to distinguish between the allocation context
argument of a function and the allocation region of its closure. For
instance, in the following example:
\begin{codejava}
letregion $\rgn$ {
  let f = $\lambda{\inang{\rho^a}}$()$\,\Rightarrow\,$new Object$\inang{\rho^a}$() 
  in f
}
\end{codejava}
The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
\ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
in $\rgn$, and (b). when executed under an allocation context
$\rho^a$, the closure returns an object allocated in $\rho^a$.

The expression language admits expressions to instantiate
($e\inang{\ralloc\rbar}$) region parameters, and to apply
($e(\bar{e})$) functions. For a method call, however, region parameter
instantiation and method application are captured in a single
expression ($e.mn\inang{\ralloc \rbar}(\ebar)$). The language also
admits a big-lambda expression, which, when used in conjunction with
the instantiation expression, serves the bureaucratic purpose of
renaming bound region variables in the type of a function closure.

The

>>>>

The expression language has been equipped with lambda abstraction
($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
expressions to define and apply anonymous functions. These expressions
are uncurried variants of the corresponding expressions from System F.
Note that FGJ formalism does not include higher-order functions, hence
the syntactic class (\C{T}) of FGJ types needs to be extended with an
(uncurried) arrow type.

Informally, \FB class definitions are extensions of FGJ class
definitions with regions. A class definition in FGJ can be lifted to
its \FB definition by (a). lifting all core types in the body of the
class (i.e, fields, constructors and methods) to their
region-annotated versions, (b). parameterizing the class over
region variables that occur free in the region-annotated body of the
class, and (c). recording any outlives constraints on the region
parameters of the class. The later two steps effectively make the
class region polymorphic (cf. \S\ref{sec:alloc-ctxt}: qualified region
polymorphism). Note that in \FB, regions, and their static identifiers
($\rgn$) are confined to the expression language. Consequently, the
region annotated types of class fields, constructors and methods do
not refer to identifiers of concrete regions, but rather their place
holders called \emph{region variables} ($\rho$). Informally, region
variables ($\rho$) are to region identifiers ($\rgn$) as type
variables ($a$) are to core types ($T$).

@@

\FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
$\RgnZ\inang{\rgn}$. However, unlike the $\rgn$ in
$\ObjZ\inang{\rgn}$, $\rgn$ in $\RgnZ\inang{\rgn}$ cannot be any
region. Recall that $\RgnZ$ objects have special semantics in \name -
they act as handlers to transferable regions.  Constructing a new
$\RgnZ$ object entails the creation of a new transferable region, and
it is in this region that the new object is allocated in. It follows
that $\rgn$ in $\RgnZ\inang{\rgn}$ should be the static identifier of
the new transferable region. But, static identifiers for transferable
regions are introduced only when such regions are opened for
allocation via \C{open} expression. What, then, should be the region
type of a \C{new Region} expression?

@@

A method definition ($d$) can be region polymorphic with respect to
(a). its allocation context (Sec.~\ref{sec:alloc-ctxt}), and (b). the
regions occuring in the region types of its arguments.  Region
parameters on the methods, like those on classes, are accompanied by
constraints ($\phi$) capturing the conditions that the parameters need
to satisfy for the method to be considered well-formed
(Fig.~\ref{fig:fb-morewfrules}). Allocation context (usually $\rho^a$
or $\rho^a_m$) is the first and inevitable region parameter of every
method in \FB. If a method is not intended to be polymorphic with
respect to its allocation context (for example, if its allocation
context needs to be same as the allocation region of its object), then
the monomorphism needs to be encoded as an explicit equality
constraint in $\phi$.

Like methods, anonymous functions can also be region-polymorphic. 
% A lambda expression defines a region-polymorphic
% multi-argument function closure parameterized over function's
% allocation context parameter. 
The angle braces in the lambda expression ($\inang{\rho^a\rhobar \,|\,
\phi}$) serve the same purpose as they do in a method definition - to
capture region parameters along with their constraints. Region
parameters also appear in the arrow type of the lambda expression 
($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn} 
\tau$) at the prenex position, similar to ML type schemes. However,
unlike ML, we don't distinguish between region types and region type
schemes; any of the $\tau$'s in the arrow type can themselves be
region-parametric arrow types. In this respect, our region type system
is more like System F's type system, which admits higher-rank
parametric polymorphism. Like System F, \FB provides a region
instantiation expression ($e\inang{\ralloc\rbar}$), and a region
generalization expression ($\Lambdaexp{\rhoalloc\rhobar \,|\,
\phi}{e}$) to instantiate and generalize region variables.

Note that a lambda expression creates a closure, which can escape the
context in which it is created. It is therefore important to keep track of
the region in which a closure is allocated in order to avoid unsafe
dereferences. The $\rgn$ annotation above the arrow in the arrow
type denotes the allocation region of the corresponding closure. Note
that it is important to distinguish between the allocation context
argument ($\rhoalloc$) of a function and the allocation region
($\rgn$) of its closure. In \name, the later corresponds to the region where
a \C{Func} object is allocated, while the former corresponds to the
region where it is applied. 
% For instance, in the following example:
% \begin{codejava}
% letregion $\rgn$ {
%   let f = $\lambda{\inang{\rho^a}}$().$\,$new Object$\inang{\rho^a}$() 
%   in f
% }
% \end{codejava}
% The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
% \ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
% in $\rgn$, and (b). when executed under an allocation context
% $\rho^a$, the closure returns an object allocated in $\rho^a$.
